### 2025-08-09 chapters/benchmark.tex
updated TOC

\chapter{Benchmarking Evaluation}
\label{chap:benchmarking}

\section{Native Execution Performance}

\section{Enclave Execution Performance}

\section{Kyber Algorithm in the Enclave: Performance Impact}

\section{Performance Parameters: CPU Cores, Memory Allocation, Cache Size}

\section{Benchmarking Results: Native vs Enclave Execution}
%\section{Performance Bottlenecks and Overheads in Enclave Execution}

>>>>

\chapter{Benchmarking Results and Observations}
\label{chap:benchmarking}

This chapter provides a detailed empirical performance evaluation of a set of benchmark workloads executed within the Keystone enclave framework. The primary objective is to quantitatively assess the overheads and limitations introduced by enclave-based execution, focusing on how system-level parameters and hardware constraints influence workload behavior under trusted execution.

The evaluation covers a mix of general-purpose and cryptographic workloads. Specifically, we utilize the Dhrystone and CoreMark benchmarks, which are standard tools for measuring integer processing performance and embedded system efficiency, respectively. Additionally, the Kyber post-quantum cryptographic algorithm is included to evaluate a compute-intensive, security-critical workload relevant for future-proof cryptographic applications. This combination allows us to characterize enclave performance across a spectrum of computational patterns and resource demands.

\section{Baseline: Native (Non-Enclave) Performance}

To establish a baseline for evaluating enclave overheads, we first executed all benchmarks natively on the simulated RISC-V platform without any enclave isolation. This phase allows us to measure the raw performance characteristics of the system and isolate the costs introduced solely by secure execution.

The Dhrystone and CoreMark benchmarks exhibited expected scaling behavior with respect to CPU cores and memory size. As CPU core count increased, throughput for these parallelizable benchmarks improved approximately linearly, reflecting efficient multi-threaded execution on the underlying hardware. Memory size impacted CoreMark more noticeably, particularly during memory-intensive phases, where increased memory reduced paging and improved cache hit rates. In contrast, Dhrystone’s smaller working set resulted in negligible sensitivity to memory size variations.

The Kyber cryptographic benchmark behaved differently. Due to its inherently sequential algorithmic structure, Kyber’s execution time remained relatively constant across changes in CPU core count. Likewise, memory size had minimal impact on Kyber performance, as its working set comfortably fit within allocated memory without inducing paging.

These native baseline results confirm the stability and predictability of the simulated system environment, providing a solid foundation for subsequent enclave overhead analysis.

\section{Enclave Execution Results and Overheads}

Execution within Keystone enclaves introduces overhead from multiple sources: memory isolation enforced by PMP, additional context switches to manage transitions between secure and non-secure execution states, and the general overhead of enforcing enclave boundaries at runtime.

Across all benchmarks, enclave execution resulted in increased execution times compared to native runs. Overhead magnitude varied by workload and system configuration but generally remained below 15\%. For parallel workloads such as CoreMark, overhead tended to increase with the number of CPU cores, likely due to increased synchronization demands and frequent enclave exit and re-entry operations during thread management. Dhrystone showed more modest overhead growth with core scaling.

Cache behavior, measured indirectly via QEMU’s cache modeling plugin, indicated modest increases in cache miss rates during enclave execution. Although this plugin provides system-wide cache metrics rather than enclave-specific data, the observed increases likely stem from PMP-related overheads such as cache flushing or restricted memory access patterns. These overheads are an expected consequence of maintaining strong memory isolation guarantees.

A critical observation during enclave execution was the enforcement and limitation of PMP resources. The fixed PMP register count on the QEMU virt platform fundamentally restricts the number of enclaves that can be concurrently instantiated. When the number of requested enclaves exceeds the 16 PMP entries, enclave creation fails. These failures are characterized by SBI call error codes and logged kernel errors indicating that no PMP registers are available. This limitation underscores the importance of considering PMP capacity in enclave orchestration and resource allocation.

\section{Workload Sensitivity Analysis (Kyber vs CoreMark vs Dhrystone)}

The Kyber post-quantum cryptographic algorithm is a representative workload for evaluating the performance impact of enclave execution on security-critical applications. Kyber consists of three primary operations—key generation, encapsulation, and decapsulation—each profiled extensively within enclaves using the Keystone timer API to capture precise cycle counts.

Execution within enclaves increased Kyber’s runtime by approximately 7–9\% compared to native execution. This overhead is modest given the cryptographic operations’ complexity and the security benefits provided by enclaves.

As with native execution, Kyber performance was insensitive to increases in CPU core count due to its inherently sequential design. Memory usage during enclave execution remained stable, with no indications of paging or contention. System-wide cache miss rates observed through QEMU remained consistent with Kyber’s memory access patterns, including phases dominated by number-theoretic transforms and polynomial arithmetic.

From a resource perspective, the requirement of one PMP entry per enclave constrains the number of parallel Kyber instances that can be simultaneously deployed. This constraint must be taken into account when designing systems requiring multiple isolated cryptographic operations.

\section{Impact of Hardware Parameters (Cores, Memory, Cache)}

To comprehensively assess how hardware configurations affect enclave performance, we systematically varied CPU core counts, memory allocations, and cache configurations.

Scaling the number of CPU cores yielded anticipated improvements for parallel workloads, with throughput gains plateauing beyond six cores due to synchronization overhead and increased enclave management complexity. Sequential workloads such as Kyber remained unaffected by core scaling.

Memory allocation experiments revealed that increasing memory size from 64MB to 2GB improved performance for memory-intensive benchmarks by reducing paging and increasing cache hit rates. However, beyond approximately 512MB, gains diminished, indicating sufficient memory provision for the workload’s working set.

Cache configuration impacts were evaluated using the QEMU cache modeling plugin. Larger cache sizes and higher associativity reduced miss rates, particularly for instruction-heavy workloads like Dhrystone. While the plugin’s system-wide scope limits insights into enclave-specific cache activity, the trends suggest that cache tuning can meaningfully influence enclave performance.

Importantly, the relationship between memory allocation and PMP usage is nuanced. Larger enclave memory regions may require multiple PMP entries if the memory cannot be described contiguously in a single PMP register, effectively reducing the total number of enclaves supported simultaneously. This interaction between memory layout and PMP capacity represents a critical consideration for enclave deployment strategies.

### 2025-08-09 chapters/discussion.tex
updated TOC

\chapter{Discussion}
\label{chap:discussion}

\section{Key Findings and Insights}

%\section{Challenges in Benchmarking Enclaves and Cryptographic Algorithms}

\section{Trade-offs Between Security and Performance in Enclaves}

\section{Implicaitons for Secure Computing and Cryptographic Applications}

>>>>

\chapter{Interpretation and Broader Implications}
\label{chap:discussion}

\section{Synthesis of Key Performance Trends}
This chapter integrates and interprets the empirical results obtained from benchmarking Keystone enclaves, offering a comprehensive synthesis of performance characteristics, limitations, and broader implications. The experimental data reveal several critical insights into the practicalities and challenges of enclave execution on RISC-V platforms, which have direct relevance to the design and deployment of secure computing systems.

First and foremost, enclave execution imposes non-negligible but moderate overheads compared to native, non-isolated execution. These overheads predominantly arise from the enforcement of memory isolation mechanisms, including Physical Memory Protection (PMP), and the additional context switches required to manage transitions between secure (enclave) and non-secure (host) execution states. Our results show that, across a diverse set of benchmarks ranging from general-purpose integer workloads to specialized cryptographic algorithms, overheads typically remain under 15%. However, this aggregate figure masks significant variability depending on workload type, degree of parallelism, and system configuration parameters such as CPU cores, memory allocation, and cache hierarchy.

A second major insight relates to the critical role of hardware-enforced resource constraints, specifically the limited number of PMP entries. The QEMU virt platform utilized in our experiments provides 16 PMP registers, which directly limits the number of isolated enclaves that can simultaneously operate. Attempts to instantiate more enclaves than PMP entries available result in immediate failure of enclave creation, as indicated by distinct Supervisor Binary Interface (SBI) error codes and kernel log messages. This finding underscores the fundamental impact of hardware architecture on enclave scalability and points to PMP capacity as a bottleneck that must be carefully managed in real-world systems.

Thirdly, workload characteristics strongly influence enclave performance overheads. Compute-bound, sequential workloads such as the Kyber post-quantum cryptographic algorithm experience comparatively low performance degradation within enclaves. This is consistent with Kyber’s minimal inter-thread communication and largely predictable memory access patterns. In contrast, highly parallelizable benchmarks like CoreMark exhibit increasing overhead with core count, reflecting the compounded costs of synchronization, enclave entry/exit transitions, and possibly contention in PMP configuration. This differential sensitivity implies that enclave performance optimization must be workload-specific, taking into account application parallelism and memory footprint.

Lastly, the interaction between resource allocation parameters—CPU cores, memory size, and cache configuration—and enclave performance is multifaceted. Increasing CPU cores yields diminishing returns beyond a threshold due to synchronization and enclave management overhead. Larger memory allocations improve performance primarily by reducing paging and improving cache locality but also can exacerbate PMP usage if memory regions are fragmented. Cache configuration impacts, though indirectly measured via QEMU’s system-wide cache model, suggest that optimizing cache size and associativity can mitigate enclave overheads, especially for instruction-intensive workloads. These observations highlight the necessity of a holistic approach to system tuning in enclave environments.

\section{Hardware Bottlenecks and PMP Utilization Strategies}
A central and recurring theme throughout this study is the pivotal role played by the RISC-V Physical Memory Protection (PMP) mechanism in determining the scalability, flexibility, and reliability of secure enclave execution in the Keystone framework. Although PMP is indispensable for maintaining strong memory isolation—arguably the most critical security property in any Trusted Execution Environment (TEE)—it simultaneously introduces significant architectural constraints that place hard limits on system scalability. These limitations become especially apparent in environments that demand the concurrent instantiation of multiple enclaves, such as secure edge computing deployments, cloud-based confidential workloads, and multi-tenant embedded systems.

PMP operates by establishing fine-grained access control policies over physical memory ranges. These policies are enforced in hardware and serve to prevent unauthorized access by less privileged software layers, such as applications running in Supervisor or User mode. In the Keystone architecture, enclaves rely on PMP to ensure that their memory regions remain inaccessible to both the host operating system and other enclaves. Each protected memory region is described by a PMP entry, and these entries are programmed through control and status registers (CSRs) accessible only from the highest privilege level—Machine mode (M-mode). Each PMP entry specifies a memory region, along with the access permissions allowed (read, write, execute), and once configured, becomes an unforgeable part of the system’s isolation model.

However, the number of available PMP entries is not dynamic—it is a fixed architectural parameter determined by the processor implementation. In our experimental setup based on the widely used QEMU \texttt{virt} platform for RISC-V, this number is statically defined as 16. This finite resource pool becomes a fundamental limiting factor in the Keystone system because \textbf{each enclave requires at least one PMP entry} to establish its isolated memory region. In more complex memory layouts—especially those that are large or fragmented—multiple PMP entries may be needed to define non-contiguous or irregularly sized memory regions, further exacerbating resource pressure.

This constraint manifested clearly in our benchmarking experiments. When the number of concurrently requested enclaves exceeded the available PMP entries, enclave creation failed immediately. These failures were not silent; they were explicitly reported by the system in the form of Supervisor Binary Interface (SBI) error codes and verbose kernel log messages, providing clear indicators of PMP exhaustion. Typical output during such events included:

\begin{lstlisting}[language=bash,caption={Error messages during enclave creation and destruction}]
napot_region_init: No available PMP register
keystone_create_enclave: SBI call failed with error code 100003
fatal: cannot destroy enclave: SBI failed with error code 100005
pmp_unset: Invalid PMP region index
pmp_region_free_atomic: Invalid PMP region index
\end{lstlisting}

These logs reflect a hard architectural ceiling that is entirely independent of other system resources such as CPU cores or DRAM availability. Even in scenarios where there is ample memory and processing capacity to support additional enclaves, the inability to allocate a free PMP entry effectively blocks the deployment of further isolated workloads. This introduces a non-obvious, hardware-enforced bottleneck that could significantly impact the reliability and scalability of enclave-enabled systems.

Moreover, the situation is further complicated by the way PMP regions are encoded. PMP supports a limited set of region encoding formats, with the most commonly used being the Naturally Aligned Power-of-Two (NAPOT) mode. This format requires that protected memory regions be both naturally aligned and sized as a power of two. While this encoding simplifies hardware enforcement, it imposes rigid constraints on memory layout. For example, if an enclave's memory region spans an irregular range—such as 192 KB starting at a non-aligned address—it cannot be expressed with a single PMP entry. In such cases, the region must be broken down into smaller, aligned sub-regions, each requiring its own PMP entry. This fragmentation effect leads to inefficient utilization of available PMP entries and further reduces the maximum number of enclaves that can coexist.

To address these challenges, both \textbf{hardware-level} and \textbf{software-level} mitigation strategies must be considered:

\begin{itemize}
    \item \textbf{Hardware approaches} include increasing the number of available PMP entries. While this would require architectural changes and potentially more silicon real estate, it could dramatically improve enclave concurrency. Alternatively, future iterations of the RISC-V PMP specification could explore more flexible memory region encodings, allowing for more expressive and compact representations of protected regions, thereby reducing fragmentation.
    \item \textbf{Software techniques} include dynamic PMP entry management, where PMP entries are reused or reconfigured at runtime based on the current workload and enclave activity. This requires careful coordination to avoid security violations but can enable more flexible use of the limited register set. Another technique is enclave multiplexing, where multiple lightweight tasks or applications are combined into a single enclave, reducing the total number of enclaves required. Additionally, memory layout optimization—aligning and sizing enclave memory regions to match NAPOT constraints—can significantly reduce the number of PMP entries consumed per enclave.
\end{itemize}

From a systems engineering perspective, these limitations have real-world consequences. In cloud environments, where tenants may each be assigned their own isolated enclave for security, PMP constraints could become a source of contention, limiting the number of clients that can be served simultaneously. In embedded or edge computing devices with strict memory constraints, careful tuning of enclave footprints and memory allocation strategies will be necessary to ensure system robustness.

In conclusion, while the PMP mechanism is undeniably a cornerstone of the Keystone security model, it also introduces a set of hard limitations that impact the practical deployment and scaling of enclaves. Understanding and addressing these limitations—both through hardware evolution and smarter software resource management—is essential for building scalable, secure computing platforms based on the RISC-V architecture.

\section{Implications for Secure System Design and Cryptographic Workloads}

\section{Recommendations and Considerations for Future TEE Deployments}
